---
title: "Binned Effect Size"
---
```{r}
library(dplyr)
library(ggplot2)
```
```{r}
unlinked_sites <-  read.csv("unlinked_sites.csv")
adult <- unlinked_sites[which(unlinked_sites$Age == "Adult"),-c(1:3)]+2
juv <- unlinked_sites[which(unlinked_sites$Age == "Juv"),-c(1:3)]+2
```
Get the allele frequencies for adults and juveniles
```{r}
allele_frequencies <- data.frame(Site = colnames(adult), JuvFreq = colSums(juv)/453, AdultFreq = colSums(adult)/129)
head(allele_frequencies)
```
```{r}
plot(allele_frequencies$JuvFreq, allele_frequencies$AdultFreq)
cor.test(allele_frequencies$JuvFreq, allele_frequencies$AdultFreq)
```


```{r}
allele_frequencies1 <- read.csv("allele_frequencies.csv")
MP_effect_sizes <- read.csv("effect_sizes.csv")
```

Join the two
```{r}
MP_effect_sizes2 <- data.frame(paste(MP_effect_sizes$CONTIG, MP_effect_sizes$SNP, sep = "_"), MP_effect_sizes$EffectSize)
colnames(MP_effect_sizes2) <- c("Site","EffectSize")
```

```{r}
effect_and_freq <-  merge(MP_effect_sizes2, allele_frequencies, by = "Site")
effect_and_freq <- data.frame(effect_and_freq,
                              Diff = (effect_and_freq$JuvFreq - effect_and_freq$AdultFreq),
                              DiffADJ = (effect_and_freq$JuvFreq - effect_and_freq$AdultFreq)/effect_and_freq$AdultFre)
```
Bin the effect sizes

```{r}
dmean <- c()
ese <- c()

bins <- 200
brks <- quantile(effect_and_freq$EffectSize, seq(0,1, length.out = bins+1))
es_mid <- (brks[-1]+brks[-(bins+1)])/2
dmean <- rep(0,bins)
ese <- rep(1, bins)
for (i in 1:bins) {
  choice <- effect_and_freq$EffectSize>brks[i] & effect_and_freq$EffectSize<brks[i+1]
  dmean[i] <- mean(effect_and_freq$Diff[choice])
  ese[i] <- sd(effect_and_freq$Diff[choice])/sqrt(sum(choice))
  }

binned_data <- data_frame(MeanEffect = es_mid, MeanChange = dmean)
```
```{r}
af_change_plot_1 <- ggplot(data = binned_data, aes(x = MeanEffect, y = MeanChange/2))+
  geom_point(shape=21, size = 2, fill = "white")+
  geom_abline()+
  theme_minimal()+
  xlab("Effect size (200 quantiles)")+
  ylab("Mean change in allele frequency")

af_change_plot_1
```

```{r}
effect_and_freq_bin <- effect_and_freq %>% mutate(new_bin = ntile(EffectSize, n=200))

res <- effect_and_freq_bin %>% group_by(new_bin) %>%
        summarise(Num = n(), MeanEffect = mean(EffectSize,na.rm=TRUE))

res2 <- effect_and_freq_bin %>% group_by(new_bin) %>%
        summarise(Num = n(), MeanChange = mean((JuvFreq-AdultFreq),na.rm=TRUE))

binned_data <- merge(res, res2, by = "new_bin")
```
```{r}
af_change_plot <- ggplot(data = binned_data, aes(x = MeanEffect, y = MeanChange/2))+
  geom_point(shape=21, size = 2, fill = "white")+
  geom_smooth(method = "lm", se=T) +
  theme_minimal()+
  xlab("Effect size (200 quantiles)")+
  ylab("Mean change in allele frequency")
 # ylim(-0.02, 0.02)+
 # ggtitle("B")

af_change_plot
```
```{r}
unbinned_plot <- ggplot(data = effect_and_freq, aes(y=Diff, x = EffectSize))+
  geom_point(pch=".")+
  #geom_smooth(method = "lm", se=T) +
  theme_minimal()+
  xlab("Effect size")+
  ylab("Change in allele frequency")
  #ylim(-0.32, 0.32)+
  #ggtitle("A")

unbinned_plot
```

```{r}
pq <- effect_and_freq$JuvFreq*(1- effect_and_freq$JuvFreq)
lm_effect <- lm(effect_and_freq$Diff~effect_and_freq$EffectSize)
summary(lm_effect)
lm_effect_freq <- lm(effect_and_freq$Diff~effect_and_freq$EffectSize*pq)
summary(lm_effect_freq)
```
```{r}
af <- (effect_and_freq$AdultAlleles+effect_and_freq$JuvAlleles)/(2*575)
es <- effect_and_freq$EffectSize

maf <- c()
for (i in 1:nrow(effect_and_freq)){
  if(af[i]>0.5){
    maf[i] <- 1-af[i]
    es[i] <- es
  }
  else{
    maf[i] <- af[i]
  }
}

pq <- maf*(1- maf)

diff <- effect_and_freq$Diff

lm1 <- lm(diff~es)
pred1 <- predict(lm1, newdata = data.frame(es = c(es)))
plot(pred1, diff, pch=".")

lm2 <- lm(diff~es*pq)
pred2 <- predict(lm2, newdata = data.frame(es = c(es), pq = c(pq)))
plot(pred2, diff, pch=".")
plot(es, pred2, pch=".")
```
`Nb I used mean allele frequency over both generations, as I wanted to capture the frequency of alleles in the missing parents.

Nb I don’t use exactly pq x es but the difference is negligible (graph 1)

To explain, if we assume the focal allele has a fitness of 1 in the homozygotes and 1-s in the hets then the change in allele frequency is 
The deficit in the heterozygotes i.e (2pqs x 1/2) / mean fitness. 

The x 1/2 is because only half the het alleles are the focal allele.  The approximation is assuming mean fitness is 1, which close to true these small fitness effects.
I used the full formulae for my analysis.

```{r}

```

```{r}
library(cowplot)
both_plots <- ggdraw()+
  draw_plot(unbinned_plot, x = 0, y = 0.5, height = 0.5)+
  draw_plot(af_change_plot, x = 0, y = 0, height = 0.5)

both_plots

both_plots2 <- ggdraw()+
  draw_plot(unbinned_plot, x = 0, y = 0, width = 0.5)+
  draw_plot(af_change_plot, x = 0.5, y = 0, width = 0.5)

both_plots2
```

